---
title: "Printing test and output from targets pipeline"
format: pdf
---

```{r setup}
#| include: false
library(knitr)
library(targets)
library(tarchetypes)
```

```{r}
tar_read(test_most_common)
tar_read(test_naturetypes_p)
temp <- tar_read(naturetypes_wide)
temp |> head()
tar_read(plot_naturetypes_comb)
```

The NA's represents localities where just one of the two variables (then thinking 7SE and PRSL as the same variable)
is recorded. To combine these into one metric, ADSV, I could take the one with the highest value (worst-rule) or the sum.
Sum is problematic as not all locations have two values to sum together.But the other option is problematic since I think field workers often tend to split the effects over two variables is they have that option. And if we have 50% vehicle damage and 50% hiking damage, that is no doubt worst than just having 50% of either. So I will use the sum, despite its issues.

```{r}
tar_read(plot_naturetypes_comb2)
```

```{r}
temp <- tar_read(naturetypes_wide2)
temp |> head()
```
