---
title: "bayesUpdating"
format: html
editor: visual
---

## Function to update an estimate of a zero and 1-inflated beta distributed distribution

The goal here is to use a larger dataset (national dataset) to infer about the shape of a smaller domain dataset. The domain mean will always be uninformed by the national data, but I will use empirical bayes to make a better guess at the true variation of the domain data when sample sizes are low. I add a prior weight of 10, meaning the national dataset coundt similarly as ten domain datapoints in describing the distributional shape for the domain data.

Deepseek (free version, may-juni 2025) was used to flesh out some of the ideas and test for logical errors. However, the original idea was by the authors, and all code has been verified and we also added a simulation to test the behaviour of the functions.

I will now simulate som (national) data and some domain data
and then test the functions for getting updated domain shape and variance.

I tried also with prior weight of 5, and results are qualitativly
similar. As a rule of thumb, I want the posterior distribution to include
just about the whole range for the national data when n = 2-3.

This example does not include area weighting. Latyer I want to use area weighting for the domain means, but I will not include it for the updated shape parameters.

```{r setup}
library(tidyverse)
library(MASS)
library(ggh4x)
```

```{r nationalData}
# set seed
set.seed(123)

# define sample size
n <- 1000


national_data <- tibble(
  # first, some population samples with different properties
  uniform = runif(n, 0, 1),
  normal_centered = rnorm(n, 0.5, 0.2),
  normal_high = rnorm(n, 1, .5),
  beta_2_2 = rbeta(n, 2, 2),
  beta_05_3 = rbeta(n, .5, 3),
  beta_3_05 = rbeta(n, 3, .5),
) |>
  # truncate
  mutate(across(
    !starts_with("wgt"),
    ~ case_when(
      . >= 1 ~ 1, 
      . <= 0 ~ 0, 
      .default = .
    )
  )) |>
  pivot_longer(
    cols = !starts_with("wgt"),
    names_to = "distribution",
    values_to = "population_sample"
  ) |>
  nest(population_sample = population_sample, wgt = wgt)

    
```

```{r domainData}


# pre set some sample sizes
a <- 2
b <- 5
c <- 10
d <- 30

# set seed
set.seed(123)

domain_data <- tibble(
    sample_ID = c(
      "domain_1a",
      "domain_1b",
      "domain_1c",
      "domain_1d",
      "domain_2a",
      "domain_2b",
      "domain_2c",
      "domain_2d",
      "domain_3a",
      "domain_3b",
      "domain_3c",
      "domain_3d",
      "domain_4b",
      "domain_4c",
      "domain_4d",
      "domain_5b",
      "domain_5c",
      "domain_5d"),
    
    values = list(
      # 1's
      tibble(sample = rep(1,a)), 
      tibble(sample = rep(1,b)),
      tibble(sample = rep(1,c)),
      tibble(sample = rep(1,d)),
      
      #  high values
      tibble(sample = seq.int(0.7, 1, length.out=a)),
      tibble(sample = seq.int(0.7, 1, length.out=b)),
      tibble(sample = seq.int(0.7, 1, length.out=c)),
      tibble(sample = seq.int(0.7, 1, length.out=d)),
      
      #  low values (further from the population shape)
      tibble(sample = seq.int(0.1, 0.4, length.out=a)),
      tibble(sample = seq.int(0.1, 0.4, length.out=b)),
      tibble(sample = seq.int(0.1, 0.4, length.out=c)),
      tibble(sample = seq.int(0.1, 0.4, length.out=d)),
      
      # truncated normal with sd = .5, which is probably more variation than the natinal data
      tibble(sample = rnorm(b, 1, .5) |> pmin(1) |> pmax(0) |> sort()),
      tibble(sample = rnorm(c, 1, .5) |> pmin(1) |> pmax(0) |> sort()),
      tibble(sample = rnorm(d, 1, .5) |> pmin(1) |> pmax(0) |> sort()),
      
      # truncated normal with sd = .1
      tibble(sample = rnorm(b, 1, .1) |> pmin(1) |> pmax(0) |> sort()),
      tibble(sample = rnorm(c, 1, .1) |> pmin(1) |> pmax(0) |> sort()),
      tibble(sample = rnorm(d, 1, .1) |> pmin(1) |> pmax(0) |> sort())
      
    ))


    
```

```{r tbl-showData}

DT::datatable(national_data)
```

```{r tbl-showData}

DT::datatable(domain_data)
```

```{r fig-plotDist}

# plot the population samples
national_data |>
  unnest(population_sample) |>
  ggplot()+
  geom_histogram(aes(x = population_sample),
                 binwidth = 0.1)+
  facet_wrap(vars(distribution),
             scales = "free")
#ggsave("../images/population_samples.png")
```

```{r fig-plotDist}

# plot the domain samples
domain_data |>
  unnest(values) |>
  ggplot()+
  geom_histogram(aes(x = sample),
                 binwidth = 0.1)+
  xlim(-0.2, 1.2)+
  facet_wrap(vars(sample_ID),
             scales = "free_y")
#ggsave("../images/population_samples.png")
```

```{r fit_zoib}
# function to fit zero and one inflated beta distributions
fit_zoib <- function(x) {
  prop_1 <- mean(x == 1)
  prop_0 <- mean(x == 0)
  x_trunc <- x[x > 0 & x < 1]
  alpha <- MASS::fitdistr(x_trunc, dbeta, start = list(shape1 = 1, shape2 = 1))$estimate
  list(alpha = alpha, prop_1 = prop_1, prop_0 = prop_0)
}


```

```{r}
# applied on the national dataset

national_data <- national_data |>
  rowwise() |>
  mutate(fit = 
           list(fit_zoib(unlist(population_sample))))

```



```{r updatingFunction}
# Empirical bayes
# Bayesian updating: combine domain data with national shape

update_distribution <- function(domain_data, national_fit, prior_weight = 10) {
  
  n <- length(domain_data)
  prop_1_domain <- mean(domain_data == 1)
  prop_0_domain <- mean(domain_data == 0)
  
  # Update 1-inflation (weighted average)
  updated_prop_1 <- (n * prop_1_domain + prior_weight * national_fit$prop_1) / (n + prior_weight)
  updated_prop_0 <- (n * prop_0_domain + prior_weight * national_fit$prop_0) / (n + prior_weight)
  
  # Update beta parameters (only for 0 < x < 1 part)
  x_trunc <- domain_data[domain_data > 0 & domain_data < 1]
  
  if (length(x_trunc) > 1) {
    tryCatch({
      alpha_domain <- MASS::fitdistr(x_trunc, dbeta, 
                                    start = list(shape1 = 1, shape2 = 1))$estimate
      
      updated_alpha <- (alpha_domain * length(x_trunc) + national_fit$alpha * prior_weight) /
                          (length(x_trunc) + prior_weight)
    }, error = function(msg){
      return(NA)
    }
      
    )
  } else {
    updated_alpha <- national_fit$alpha
  }
  
  if(!exists("updated_alpha")) updated_alpha <- NA
  
  list(
    alpha = updated_alpha, 
    prop_1 = updated_prop_1, 
    prop_0 = updated_prop_0
    )
}

#updated_dist <- update_distribution(domain_data, national_fit)
```

```{r}
# applied on the domain samples

comb_data <- domain_data |>
  cross_join(
    national_data)

comb_data <- comb_data |>
  rowwise() |>
  mutate(updated_fit = 
           list(
             update_distribution(
               domain_data = unlist(values),
               national_fit = fit
           )))
```

```{r simulationFunction}

# Generate uncertainty distribution around domain mean
simulate_domain <- function(dist, n_sim = 10000) {
  is_1 <- rbinom(n_sim, 1, dist$prop_1)
  is_0 <- rbinom(n_sim, 1, dist$prop_0) * (1 - is_1)
  cont_part <- (1 - is_1 - is_0) * rbeta(n_sim, dist$alpha[1], dist$alpha[2])
  is_1 * 1 + is_0 * 0 + cont_part
}

```

```{r}
# Applied on the test cases
comb_data <- comb_data |>
  rowwise() |>
  mutate(PDs = 
    list(simulate_domain(updated_fit, n_sim = 1000)))

```

```{r}
# add sample means (unweighted in this case)

comb_data <- comb_data |>
  mutate(
    mean = mean(unlist(values)),
    firstQuartile = quantile(unlist(PDs), 0.25),
    thirdQuartile = quantile(unlist(PDs), 0.75),
    lowCI = quantile(unlist(PDs), 0.025),
    highCI = quantile(unlist(PDs), 0.975)
  )
```


```{r}
# Prepare data for plotting
comb_data <- comb_data |>
  mutate(
    n = str_sub(sample_ID, -1, -1),
    n = case_when(
      n == "a" ~ 2,
      n == "b" ~ 5, 
      n == "c" ~ 10,
      n == "d" ~30
    ),
    scenario = str_sub(
      sample_ID, -2, -2
    ),
    scenario = case_when(
      scenario == "1" ~ "Only 1's",
      scenario == "2" ~ "High values",
      scenario == "3" ~ "Low values",
      scenario == "4" ~ "Trunc.nor. high SD",
      scenario == "5" ~ "Trunc.nor. low SD"
    )
  )
```

```{r fig-bigFacet}
#| fig-height: 12
#| warning: false

comb_data |>
  ggplot()+
  geom_density(
    data = comb_data |>
      unnest(population_sample),
    aes(x = population_sample,
        y = after_stat(scaled)),
    fill = "sienna1",
    alpha = .5
  ) +
  geom_density(
    data = comb_data |>
      unnest(PDs),
    aes(x = PDs,
        y = after_stat(scaled)),
    alpha = .5,
    fill = "cornflowerblue"
  ) +
  geom_histogram(
    data = comb_data |>
      unnest(values),
    aes(x = sample,
        y = stat(ncount)),
    binwidth = .05,
    alpha=.8,
    fill = "cornflowerblue"
  )+
      theme(legend.position = 'none',
        axis.title.y = element_blank()) +
  geom_segment(
    aes(x = mean,
        xend = mean,
        y = 0,
        yend = .2),
    linewidth = 3,
    alpha=.7,
    colour = "grey10"    
  )+
  geom_segment(
    aes(x = lowCI,
        xend = highCI,
        y = -0.1,
        yend = -0.1),
    linewidth = 1.2,
    alpha=.7,
    colour = "grey30"    
  )+
  geom_segment(
    aes(x = firstQuartile,
        xend = thirdQuartile,
        y = -0.1,
        yend = -0.1),
    linewidth = 3,
    alpha=.7,
    colour = "grey30"    
  )+
  xlim(-0.2,1.2)+
  facet_nested(scenario + n ~distribution)

ggsave("../images/posteriorCheck.png",
       width = 30, height = 30, units = "cm")

```



